Evaluating the Performance and Energy Efficiency of the COSMO-ART model system

### 2014-04-24 trunk/paper_enahpc14/Sec3:Environment-setup.tex
reduces size of paper to 7 pages

We have  chosen a state-of-the-art Intel's  third-generation Core (aka
Ivy  Bridge) processing platform  at CSCS  (called ``Monch'')  for our
power  measurements,  which  is  slated  to stay  in  service  without
hardware  upgrade  for the  duration  of  the  Exa2Green project.   In
principle at least,  this architecture could be recreated  or found in
an identical configuration beyond  the lifetime of the project.  Since
the baseline benchmark can  be reproduced within an expected variance,
and  that the baseline  run configuration  can be  used in  all future
versions  of the  code, a  fair comparison  will be  made  between the
baseline and  the milestone versions  of COSMO-ART.  In this  study, a
complementary  energy-to-solution benchmarking  comparison  is carried
out   on  the   ``Pilatus''  cluster   at  CSCS,   based   on  Intel's
previous-generation  Sandy  Bridge  processors,  conventional  in  HPC
systems and known to be  more power consuming.  Finally a full tracing
experiment is  conducted on the  ``Tintorrum'' cluster at the  Jaume I
University (UJI) using an  integrated framework, to capture an overall
power profile at a much finer resolution and to get more insights into
the MPI blocking and polling influences on power savings.

\subsubsection{CSCS - Monch}
Monch is  a 10 rack  NEC-provided and dual-socket Intel  Ivy Bridge-EP
based  cluster, utilised  by  scientists  who are  part  of the  Swiss
Platform for Advanced Scientific  Computing (PASC) \citep{PASC}. It is
composed of 312 standard  compute nodes, 24 large-memory compute nodes
and  24  huge-memory  compute   nodes.   Each  standard  compute  node
comprises two  Intel Ivy Bridge Efficient Performance  (EP) E5-2660 v2
ten-core processors operating at  2.2 GHz base clock speed, themselves
connected by a high speed  InfiniBand network based on Mellanox SX6036
managed  FDR switches,  with a  56 Gb/s  speed.  Each  core has  32 KB
instruction and 32  KB data L1 caches  and 256 KB of L2  cache. All 10
cores share a  25 MB L3 cache  and the platform has 32GB  of DDR3 1600
MHz RAM.  For our energy-to-solution  benchmark, a full rack  of Monch
constituted  of  52   standard  compute  nodes  (monchc[029-080])  was
considered.

\subsubsection{CSCS - Pilatus} 
Pilatus  is  a dual-socket  eight-core  Intel  Sandy  Bridge EP  based
cluster used as Piz Daint pre-post processing cluster.  It is composed
of 42 compute  nodes and has 2 high-speed  interconnects based on FDR:
the  first is  dedicated to  the  MPI traffic  and the  second to  the
storage high  speed traffic.   The 2 login  nodes and the  42 computes
nodes  consists in  11  twin-pair Intel  E5-Series  DALCO r2264i4t  2U
scalable compute modules.  Each  module contains 4 compute nodes based
on two Intel  Xeon E5-2670 processors operating at  2.6 GHz base clock
speed, themselves  connected by a high speed  InfiniBand network based
on Mellanox  SX6036 managed FDR switches,  with a 56  Gb/s speed. Each
core has 32 KB  instruction and 32 KB data L1 caches  and 256 KB of L2
cache.  All  the 8 core share  a 20 MB  L3 cache and the  platform has
64GB of  DDR3 1600 MHz  RAM.  For our energy-to-solution  benchmark, a
full  rack  of  Pilatus  constituted  of  42  standard  compute  nodes
(pilatus[03-44]) was considered.

\subsubsection{UJI - Tintorrum} 
Tintorrum  is a heterogeneous  cluster composed  of 28  compute nodes.
For our experiments only a subset of these nodes were considered. This
set is  composed of 16  nodes, each of  which includes two  Intel Xeon
E5645 hexa-core  processors at 2.40  GHz connected via  Infiniband QDR
(Mellanox MTS3600 switch).  Each core  has 32 KB instruction and 32 KB
data L1 caches and 256 KB L2 cache. The 6 cores share a 12 MB L3 cache
and the platform has 288 GB of DDR3 1333 MHz.

@@

\subsubsection{CSCS - E3METER metering products}
Supercomputer  clusters considered  for our  experiments at  the Swiss
National Supercomputing Center (CSCS)  of ETH Zurich are equipped with
E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
high   quality   electricity  meters   released   by  Riedo   Networks
\citep{Riedonetworks},   that  enable   to  monitor   and   log  power
consumption  of the IT  infrastructure as  well as  constantly analyze
line  voltage,  current, power-factor,  frequency  with 1\%  accuracy.
Using  reliable narrowband  powerline communication  (PLC) technology,
all  metering and  power  quality  data from  each  IPS are  centrally
collected  by the E3METER  Data Concentrator,  via the  existing power
cables thus  avoiding the need for  extra cabling.  This  data is made
available via  SNMP, HTTP, TELNET  through the built-in  Fast Ethernet
port.   Time  synchronisation  is  guaranteed by  using  NTP  servers.
Measured  data is  accessed  through the  open  source Cacti  software
including the E3METER Cacti Plugin  to scan the entire PLC network and
monitor in real-time the power usage of individual rack, recorded in 5
minutes interval periods.

\subsubsection{UHAM - UJI}
To assess the  performance and the energy efficiency  of COSMO-ART, we
employ   a    version   of   the    integrated   framework   presented
in~\cite{energy13} that works in  combination with Extrae and Paraver,
which are profiling/tracing and visualization tools, respectively.

%The left part of the \vref{fig:Lustre} offers a graphical representation of
%the Lustre architecture; the right depicts the tracing and profiling framework.
To use our  approach, COSMO-ART is compiled using  the Extrae compiler
wrappers,  which  automatically instrument  the  Fortran  code of  the
model. Next, COSMO-ART  is run on the nodes,  thus dissipating certain
amount  of power.   These  nodes are  connected  to power  measurement
devices that account for the dissipated power/consumed energy and send
the power  data to the  tracing server.  The client,  meanwhile, sends
start/stop  primitives  in  order  to  gather  captured  data  by  the
watt-meters onto the  tracing server, where an instance  of the \pmlib
server is running.

Once COSMO-ART run is finished, a file containing the power profile is
created using the power data  received from the tracing server and the
instrumentation post-processing generates the performance trace files.
All  these files  are  combined then  in  Paraver which  allows us  to
visualise the performance trace and the power profile of COSMO-ART all
together.

@@

The COSMO-ART baseline  is a pure MPI based  Fortran~90 code currently
running on distributed multi-core systems only.  The software stack on
both CSCS  platforms was controlled using the  modules framework which
gives an easy and flexible mechanism  to access to all of the provided
compilers, tools  and applications.  For our  initial benchmarking, we
opted for the GNU compiler  (gcc/4.8.1 on Monch, gcc/4.8.2 on Pilatus)
using the -O3  compiler flag in favor of  the intel compiler (14.0.1),
delivering inferior performance.  In addition, we installed the MPICH2
implementation of MPI (mvapich2/1.9) as well as the commonly used HDF5
(1.8.12) and NetCDF (4.3.1)  libraries for the management of extremely
large  and  complex data  collections.   All  computes  nodes have  an
operating      system      based      on      GNU/Linux      featuring
``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
``3.0.101-0.15-default'' kernel in Pilatus.

\subsubsection{Run configuration}
A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
Pilatus.   The  calculated  region  was mapped  to  the  participating
processors using  a 2D-partitioning strategy.   The distribution along
the $x$  and $y$ coordinates  was defined by setting:  $nprocx=40$ and
$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
usually  kept  bigger than  $nprocy$.   Besides,  as  this version  of
COSMO-ART  doesn't  make  use  of  the traditional  GRIB  library,  we
specified $nprocio=0$ for GRIB  I/O.  Hyperthreading is not considered
in this study as previous attempts  of its use revealed that it always
led to higher energy-to-solution.\\

>>>>

Benchmarks presented  in this work  were performed on  three different
platforms  and can  be reproduced  within an  expected  variance, thus
allowing a  fair comparison between the baseline  and future milestone
versions of COSMO-ART.

\begin{itemize}
\item \textsc{Monch (CSCS)} is a 10 rack NEC-provided cluster composed
  of 312 compute  nodes and utilised by members  of the Swiss Platform
  for Advanced Scientific  Computing (PASC) project \citep{PASC}. Each
  node comprises  two Intel Ivy Bridge EP  E5-2660 ten-core processors
  operating at  2.2 GHz connected  via InfiniBand Mellanox  SX6036 and
  FDR switches (56 Gb/s speed), and is equipped with 32GB of DDR3 1600
  MHz  RAM.   For our  experiments,  a  full  rack constituted  of  52
  standard compute  nodes was used.  \textsc{Monch} is  slated to stay
  in  service  without  hardware  upgrade  for  the  duration  of  the
  Exa2Green  project,  allowing   an  identical  configuration  to  be
  recreated for future assessments of the baseline.\\

\item \textsc{Pilatus (CSCS)} is  a dual-socket eight-core Intel Sandy
  Bridge EP based cluster composed of 42 compute nodes and used as Piz
  Daint pre-post  processing cluster at  CSCS.  The 2 login  nodes and
  the 42 computes nodes consists in 11 twin-pair Intel E5-Series DALCO
  r2264i4t 2U scalable compute modules. Each module contains 4 compute
  nodes based  on two Intel  Xeon E5-2670 processors operating  at 2.6
  GHz connected  via InfiniBand Mellanox  SX6036 and FDR  switches (56
  Gb/s speed), and is equipped with 64GB of DDR3 1600 MHz RAM. For our
  experiments, a  full rack constituted  of 42 standard  compute nodes
  was  used. \textsc{Pilatus}  is based  on  Intel's second-generation
  processors, conventional in  HPC systems and known to  be more power
  consuming than Ivy Bridge successor.\\

\item \textsc{Tintorrum (UJI)} is  a heterogeneous cluster composed of
  28 compute nodes.  For our  experiments only a subset of these nodes
  were considered.   This set is composed  of 16 nodes,  each of which
  includes  two Intel  Xeon  E5645 hexa-core  processors  at 2.40  GHz
  connected via  Infiniband QDR (Mellanox MTS3600  switch).  Each core
  has  32 KB  instruction  and 32  KB data  L1  caches and  256 KB  L2
  cache. The 6 cores  share a 12 MB L3 cache and  the platform has 288
  GB  of DDR3 1333  MHz.  A  full tracing  experiment is  conducted on
  \textsc{Tintorrum}, to  capture an overall  power profile at  a much
  finer resolution, and to get more insights into the MPI blocking and
  polling influences on power savings.
\end{itemize}

@@

\subsubsection{CSCS - E3METER metering products}

Supercomputer  clusters considered  for our  experiments at  the Swiss
National Supercomputing Center (CSCS)  of ETH Zurich are equipped with
E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
high   quality   electricity  meters   released   by  Riedo   Networks
\citep{Riedonetworks},   that  enable   to  monitor   and   log  power
consumption  of the IT  infrastructure as  well as  constantly analyze
line  voltage,  current, power-factor,  frequency  with 1\%  accuracy.
Using  reliable narrowband  powerline communication  (PLC) technology,
all  metering and  power  quality  data from  each  IPS are  centrally
collected  by the E3METER  Data Concentrator,  via the  existing power
cables thus  avoiding the need for  extra cabling.  This  data is made
available via  SNMP, HTTP, TELNET  through the built-in  Fast Ethernet
port.   Time  synchronisation  is  guaranteed by  using  NTP  servers.
Measured  data is  accessed  through the  open  source Cacti  software
including the E3METER Cacti Plugin  to scan the entire PLC network and
monitor in real-time the power usage of individual rack, recorded in 5
minutes interval periods.

\subsubsection{UHAM - UJI}

For our experiments at Jaume I University (UJI), we employed a version
of the integrated framework presented in~\cite{energy13} that works in
combination with  Extrae and Paraver, which  are profiling/tracing and
visualization tools, respectively.

%The left part of the \vref{fig:Lustre} offers a graphical representation of
%the Lustre architecture; the right depicts the tracing and profiling framework.
COSMO-ART  is  compiled  using  the Extrae  compiler  wrappers,  which
automatically  instrument  the  Fortran   code  of  the  model.  Next,
COSMO-ART  is run  on the  nodes, thus  dissipating certain  amount of
power.  These  nodes are connected  to power measurement  devices that
account for  the dissipated power/consumed  energy and send  the power
data to  the tracing server.  The client,  meanwhile, sends start/stop
primitives in  order to gather  captured data by the  watt-meters onto
the tracing server, where an instance of the \pmlib server is running.

Once  the run  is finished,  a file  containing the  power  profile is
created using the power data  received from the tracing server and the
instrumentation post-processing generates the performance trace files.
All  these files  are  combined then  in  Paraver which  allows us  to
visualise the performance trace and the power profile of COSMO-ART all
together.

@@

The COSMO-ART baseline  is a pure MPI based  Fortran~90 code currently
running on distributed multi-core systems only.  The software stack on
both CSCS  platforms was controlled using the  modules framework which
gives an easy and flexible mechanism  to access to all of the provided
compilers, tools  and applications.  For our  initial benchmarking, we
opted for the GNU  compiler (gcc/4.8.1 on \textsc{Monch}, gcc/4.8.2 on
\textsc{Pilatus}) using  the -O3 compiler  flag in favor of  the intel
compiler (14.0.1),  delivering inferior performance.   In addition, we
installed the  MPICH2 implementation of MPI (mvapich2/1.9)  as well as
the commonly used  HDF5 (1.8.12) and NetCDF (4.3.1)  libraries for the
management  of  extremely large  and  complex  data collections.   All
computes nodes  have an operating system based  on GNU/Linux featuring
``2.6.32-358.11.1.el6.x86\_64''    kernel   in    \textsc{Monch}   and
``3.0.101-0.15-default'' kernel in \textsc{Pilatus}.

A snapshot of the code,  which includes at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was produced and run on a 1040 cores using 20~MPI tasks per
node on \textsc{Monch} and on a 1344 cores using 16~MPI tasks per node
on  \textsc{Pilatus}.   The  calculated   region  was  mapped  to  the
participating  processors  using   a  2D-partitioning  strategy.   The
distribution along the $x$ and $y$ coordinates was defined by setting:
$nprocx=40$  and $nprocy=26$  for \textsc{Monch}  and  $nprocx=28$ and
$nprocy=24$  as  $nprocx$  is   usually  kept  bigger  than  $nprocy$.
Besides,  as this  version doesn't  make use  of the  traditional GRIB
library, we  specified $nprocio=0$  for GRIB I/O.   Hyperthreading was
not considered in this study  as it previously revealed that it always
led to higher energy-to-solution.
